Ferromagnetism and spin excitations in topological Hubbard models with a flatband

### 2018-09-18 paper/paper.tex
Version 1.3

\begin{document}
\title{Ferromagnetism and spin excitations in topological Hubbard models with a flat-band}
\author{Xiao-Fei Su}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{School of Physics and Electronic Information, Huaibei Normal University, Huaibei 235000, China}
\author{Zhao-Long Gu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Zhao-Yang Dong}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Shun-Li Yu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\author{Jian-Xin Li}
\email[]{jxli@nju.edu.cn}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\date{\today}

\begin{abstract}
\par We study the spin-1 excitation spectra of the flatband ferromagnetic phases in topological Hubbard models. As a paradigm, we consider a quarter filled square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). By using the numerical exact diagonalization method with a projection onto the lower nearly-flat electron band, we determine the critical Hubbard interaction upon which the ferromagnetic phase is stable, and obtain the ferromagnetic spin-1 excitation spectra of the Chern Hubbard model and the $Z_2$ Hubbard model. Both spectra consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, in both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. We elaborate that the destabilization of the ferromagnetic phase happens when the energy of the induced roton-like modes becomes zero with the increase of the nonflatness.
\end{abstract}
\maketitle

\section{Introduction}
\par Electronic bands with nonzero topological indices reside on the center of a substantial amount of topological phenomena in condensed matter physics \cite{HK_RMP2010,QZ_RMP2011}. It was proposed in a pioneering work by Haldane \cite{H_PRL1988} that a spinless fermionic model on a honeycomb lattice exhibits integer quantum Hall effect \cite{KDP_PRL1980} without an external magnetic field. This model, serving as the first example of Chern insulator, breaks the time reversal symmetry with a complex next-nearest-neighbor hopping and is characterized by a nonzero Chern number \cite{TKNN_PRL1982}. Later, the concept was generalized to time reversal symmetric systems with spin-orbital coupling (SOC), such as the monolayer graphene \cite{KM_PRL2005a,KM_PRL2005b} and HgTe/CdTe quantum wells \cite{BHZ_S2006,Ketc_S2007}. The SOC there generates complex hopping terms similar to that proposed by Haldane but with opposite chiralities for electrons with up spins and down spins, resulting in the quantum spin Hall insulator characterized by a $Z_2$ index.

@@

\par Another involved intriguing phenomenon arising from Coulomb repulsions in flat or nearly-flat bands is the itinerant ferromagnetism \cite{T_PTP1998}. It was proved by Tasaki and Mielke that the ground state of an flat electron band with its filling factor not more than but sufficiently close to $1/2$ is ferromagnetically ordered as long as an infinitesimal onsite Hubbard repulsion is present \cite{T_PRL1992,M_PLA1993,MT_CMP1993}. Afterwards this ferromagnetism was shown to be stable against small nonflatness of the electron bands if and only if the Hubbard interaction exceeds a critical value \cite{T_PRL1994}. Spin wave excitations over this ferromagnetic ground state was also studied \cite{KA_PRL1994,SGDL_PRB2018} and itinerant topological magnons has been reported quite recently \cite{SGDL_PRB2018}.

\par The interplay between flatband ferromagnetism and nontrivial band topology enriches the related physics. In fact, ferromagnetism is essential in the generation of stable fractional Chern insulators in the proposals where the spin degrees of freedom of electrons are considered \cite{TMW_PRL2011,NSRCM_PRB2011,LBFL_PRL2012}. Furthermore, ferromagnetism can also lead to possible high-temperature quantum anomalous Hall effect (QAHE) when the nearly-flat topological band is half-filled \cite{NSRCM_PRL2012}. In this paper, we focus on this ferromagnetism induced QAHE in nearly-flat topological bands. As a paradigm, we consider a square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). When the model is quarter filled (or correspondingly, the lower nearly-flat band is half filled), the ground state is spin fully polarized due to the ferromagnetism and exhibits QAHE because of the nonzero Chern number of a single-spin band. Then the low energy physics is dominated by the one-spin-flip excitations. These spin-1 excitations have been studied by a generalized bosonization scheme where the interacting fermionic model is mapped to a free bosonic model describing spin-wave excitations at the harmonic approximation \cite{DG_PRB2015}. The ferromagnetism was shown to be stable against such spin wave excitations, which are gapless in the Chern Hubbard model and gapped in the $Z_2$ Hubbard model. However in this bosonization scheme, the free part of the electron model plays no role in the spin wave excitations other than contributing a global constant, which means it should fail due to the competition between the kinetic energy and potential energy of electrons when the nonflatness of the electron bands is not negligible \cite{T_PRL1994}. In fact, in a strictly local periodic tight-binding model, an energy band with a nonzero Chern number cannot be exactly flat \cite{CMST_JPMT2014}. Therefore it remains an open question on whether the ground state is stable against the spin-1 excitations and how the nonflatness of the electron bands manifests itself in the spin-1 excitation spectra in such models.

\par To elucidate these questions, we adopt the numerical exact diagonalization method with a projection onto the lower nearly-flat band to take close investigations on the spin-1 excitations of the models. A critical magnitude of the Hubbard interaction is found for both the Chern Hubbard model and the $Z_2$ Hubbard model, upon which the ferromagnetic phase is stable. Furthermore, the spin-1 excitation spectra are shown to consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which leads to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase. We also make a comparison of our results in the flatband limit with those obtained by the bosonization scheme \cite{DG_PRB2015}. Qualitative agreements are observed for the Chern Hubbard model, yet notable discrepancy appears for the $Z_2$ Hubbard model. We attribute the discrepancy to the self-interactions of the magnons, which is ignored in the bosonization scheme but treated exactly in our method after the projection.

\par The rest of the paper is organized as follows. In Sec. \ref{mm}, we introduce the Chern Hubbard model and $Z_2$ Hubbard model studied in this paper, discuss the nontrivial band topology of their free parts, interpret the emergence of QAHE resulting from the interplay of flat-band ferromagnetism and nontrivial band topology, and formulate the exact diagonalization method with a projection onto the lower nearly-flat band on details. In Sec. \ref{nr}, we summarize the phase diagram and elaborate the spin-1 excitation spectra of both models. Section \ref{sd} provides a summary and discussion.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{lattice}
\caption{(Color online) (a) Schematic representation of the $\pi$-flux model. Blue and red solid circles denote the $A$ and $B$ sublattices, respectively. The nearest-neighbor hopping amplitudes (solid black lines) are equal to $t_{1}\exp[i\alpha(\sigma)\pi/4]$ ($\alpha(\sigma)=1$ for the time-reversal-symmetry-breaking case and $\alpha(\sigma)=\pm1$ for the time-reversal-symmetry-preserving case) along the direction of the arrows, the next-nearest-neighbor hopping amplitudes are equal to $t_{2}$ (dashed green lines) and $-t_{2}$ (dashed purple lines). The shaded area denotes the unit cell. (b) First Brillouin Zone. $\Gamma=(0,0)$, $X_{1,2}=(\pm\pi,0)$, $Y_{1,2}=(0,\pm\pi)$, $M_{1,2,3,4}=(\pm\pi,\pm\pi)$.}
\label{lattice}
\end{figure}

\par We consider a generalized $\pi$-flux Hubbard model on the square lattice, whose Hamiltonian can be written as $H=H_0+H_U$, where $H_0$ is the generalization with electron spin of the original spinless model proposed in Ref. \cite{NSCM_PRL2011},
\begin{equation}
H_0=\sum_{\langle ij\rangle,\sigma}(t_1^{ij,\sigma}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.})
    +\sum_{\langle\langle ij\rangle\rangle,\sigma}(t_2^{ij}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.}),
\end{equation}
and $H_U$ is the Hubbard interaction
\begin{equation}
H_U=U\sum_{i}n_{i\uparrow}n_{i\downarrow}.
\end{equation}
Here, $c_{i\sigma}^{\dagger}(c_{i\sigma})$ creates (annihilates) a spin $\sigma$ electron at site $i$, $n_{i\sigma}$ is the spin $\sigma$ electron number operator at site $i$, $\langle ij\rangle$ denotes the nearest-neighbor (NN) bonds and $\langle\langle ij\rangle\rangle$ denotes the next-nearest-neighbor (NNN) bonds. As is shown in Fig. \ref{lattice}(a), the spin-dependent NN hopping amplitude $t_1^{ij,\sigma}$ and the spin-independent NNN hopping amplitude $t_2^{ij}$ are given by
\begin{equation}
t_1^{ij,\sigma}=t_1\exp\left[i\delta^1_{ij}\alpha_\sigma\pi/4\right],
\end{equation}
\begin{equation}
t_2^{ij}=t_2\delta^2_{ij}.
\end{equation}
Here, $\delta^1_{ij}=+1$ if the NN electron hopping is along the direction of the solid black arrow and $\delta^1_{ij}=-1$ if along the reversed direction. $\delta^2_{ij}=+1$ if the NNN electron hopping is along the dashed green lines and $\delta^2_{ij}=-1$ if along the dashed purple lines. The spin-dependent phase $\alpha_\sigma$ breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=\alpha_\downarrow=+1$, whereas it preserves the time-reversal symmetry but breaks the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=+1$ and $\alpha_\downarrow=-1$.

\par Due to the complex NN hopping, each electron will acquire a $\pi$ phase as it hops around a plaquette along the direction of the black arrows as indicated in Fig. \ref{lattice}(a). Therefore, $H_0$ describes free electrons hopping on a square lattice in the presence of a fictitious staggered $\pi$-flux pattern \cite{WWZ_PRB1989}. For the time-reversal-symmetry-breaking case, $\alpha_\uparrow=\alpha_\downarrow$, the flux experienced by spin-up electrons and spin-down electrons is the same, while for the time-reversal-symmetry-preserving case, $\alpha_\uparrow=-\alpha_\downarrow$, it is opposite.

\subsection{Topology of free part}\label{topo_free_part}
\par Gapped noninteracting fermionic systems can be topologically classified in the presence of symmetries by their Hamiltonians in the momentum space\cite{SRFL_PRB2008}. After the Fourier transformation, the free part $H_0$ of our model reads

@@

\par When there is a gap between the $d$ band and the $f$ band, these bands can be shown to be topologically non-trivial by calculating their Chern numbers \cite{TKNN_PRL1982} (for the time-reversal-symmetry-breaking case) or $Z_2$ indices \cite{KM_PRL2005b,SWSH_PRL2006} (for the time-reversal-symmetry-preserving case). The Chern number for a single spin component of the $d$ band or the $f$ band can be expressed in terms of the coefficients $D_{i,\mathbf{k}}$ \cite{HK_RMP2010,QZ_RMP2011},

@@

of the $d$ band is $1$ and nontrivial. As a consequence, the the ground state of $H_0$ will be a noninteracting $Z_2$ insulator and exhibits quantum spin Hall effect(QSHE) when the lower $d$ band is fully filled.

\subsection{Emergence of QAHE in half-filled nearly-flat topological bands}\label{emerg_qahe}
\par For a free fermionic system hosting an energy band with a nonzero Chern number or $Z_2$ index, the distinguished phenomenon resulting from this nontrivial band topology, such as QAHE or QSHE, only manifests itself when the topological band is fully filled. At any fractional filling, the ground state of such a system will be a trivial metal. Intriguingly, when the Coulomb interactions between electrons step in, the physics of the nontrivial band topology becomes more involved, especially when the band is nearly flat so that the effects of the Coulomb interactions are highly enhanced. Combined with strong Coulomb interactions, nontrivial topological phases can emerge from fractionally filled topological bands. In this article, we are interested in half-filled strongly-correlated nearly-flat topological bands where QAHE can arise\cite{NSCM_PRL2011,DG_PRB2015}. The essence for the occurrence of this nontrivial phase is the emergence of itinerant ferromagnetism on nearly-flat bands\cite{T_PTP1998}, which fully polarizes all the electron spins. Therefore only one spin component of the topological band will be fully filled exactly, which leads to QAHE due to the nonzero Chern number of that spin component of the band.

\par The Chern Hubbard model and $Z_2$ Hubbard model on square lattice described above serve as the paradigm, where half filling of the lower electron band corresponds to quarter filling of the whole system because of the existence of $AB$ sublattices. When $t_2/t_1$ takes values in a selected region, the $d$ band and $f$ band are quite flat in that the flatness ratio $\Delta/W$, which is defined as the ratio between the electron gap $\Delta$ between these two bands and the bandwidth $W$ of the lower band, can be as large as $4.83$ \cite{NSCM_PRL2011,DG_PRB2015}. In Ref. \cite{NSCM_PRL2011}, the quarter filled $Z_2$ Hubbard model was studied and all spin excitations were shown to be gapped in the flatband limit, which is essential for a possible high-temperature realization of QAHE. In Ref. \cite{DG_PRB2015}, both models at quarter filling were considered and a generalized bosonization scheme was developed to study the spin-1 excitation spectra. However, in their bosonization scheme, the free part $H_0$ of the Hamiltonian only contributes a constant to the spectra, thus it cannot capture the physics caused by the nonflatness of the topological electron bands, which is unavoidable because an energy band with a nonzero Chern number in a strictly local periodic tight binding model cannot be exactly flat \cite{CMST_JPMT2014}. The spin-1 excitations, as the dominant low-energy excitations, deserves closer investigations to understand the stability of such phases and the physics related to the nonflatness of the topological bands. In the next subsection, we will introduce the exact diagonalization method with a projection onto the lower nearly-flat electron band to elucidate these questions.

\subsection{Exact diagonalization with projection}\label{exact_diag_proj}
\par Exact diagonalization method with a projection onto the low-energy Hilbert space bas been widely applied to systems that host flat or nearly-flat energy bands \cite{NSCM_PRL2011,RB_PRX2011,NSRCM_PRB2011,SGDL_PRB2018,NSRCM_PRL2012,DG_PRB2015}. This approach ignores the band mixing of electrons and applies when the Coulomb interaction is small compared to the energy gap of the electron bands.

@@

Let $|\text{FM}\rangle$ denote the spin-up fully polarized state on the $d$ band,

@@

where FBZ denotes the first Brillouin zone [see Fig. \ref{lattice}(b)] and $|0\rangle$ is the fermion vaccum. Then the basis of the spin-1 excitations with a center-of-mass momentum $\mathbf{q}$ over this reference state can be written as

@@

|\mathbf{k}_i\rangle_\mathbf{q}=d^{\dagger}_{\mathbf{k}_{i}-\mathbf{q}\downarrow}d_{\mathbf{k}_{i}\uparrow}|\text{FM}\rangle.

@@

Here $d^{\dagger}_{\mathbf{k}_{i}-\mathbf{q}\downarrow}$ creates a spin-down electron with momentum $\mathbf{k}_{i}-\mathbf{q}$ and $d_{\mathbf{k}_{i}\uparrow}$ annihilates a spin-up electron with momentum $\mathbf{k}_{i}$. Therefore this basis labels a spin-1 scattering channel with the index $\mathbf{k}_{i}$. Thus the dimension of this Hilbert space scales linearly with respect to the number of electron momentums \cite{SGDL_PRB2018}, and a much larger system can be numerically accessed than the usual exact diagonalization without projection, which enables us to analyze the properties of the spin-1 excitation spectra at the whole first Brillouin zone rather than some restricted discrete points solely. The matrix element of the projected Hamiltonian on this spin-1 excitation basis can be easily obtained after some algebra,

@@

Here $\delta_{\mathbf{k}_j,\mathbf{k}_i}$ is the Kronecker delta function. Then the full spin-1 excitation spectra can be obtained by the diagonalization of the matrix whose elements are defined by Eq. (\ref{PHP}). It is noted that $|\text{FM}\rangle$ is the true ground state only if the whole spin-1 excitation spectra have no negative eigen energies. Thus we can use this as the criterion to determine the destabilization point of the ferromagnetic phase.
\par We also want to give some remarks on the flatband limit in the framework of this method. The free part $h^\text{flat}_{\mathbf{k}\sigma}$ of the Hamiltonian in the flatband limit is defined as

>>>>

\begin{document}
\title{Ferromagnetism and spin excitations in topological Hubbard models with a flatband}
\author{Xiao-Fei Su}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{School of Physics and Electronic Information, Huaibei Normal University, Huaibei 235000, China}
\author{Zhao-Long Gu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Zhao-Yang Dong}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Shun-Li Yu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\author{Jian-Xin Li}
\email[]{jxli@nju.edu.cn}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\date{\today}

\begin{abstract}
\par We study the spin-1 excitation spectra of the flatband ferromagnetic phases in topological Hubbard models. As a paradigm, we consider a quarter filled square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). By using the numerical exact diagonalization method with a projection onto the lower nearly-flat electron band, we determine the critical Hubbard interaction below which the ferromagnetic phase is unstable, and obtain the ferromagnetic spin-1 excitation spectra of the Chern Hubbard model and the $Z_2$ Hubbard model. Both spectra consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, in both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. We elaborate that the destabilization of the ferromagnetic phase happens when the energy of the induced roton-like modes becomes zero with the increase of the nonflatness.
\end{abstract}
\maketitle

\section{Introduction}
\par Electronic bands with nonzero topological indices reside on the center of a substantial amount of topological phenomena in condensed matter physics \cite{HK_RMP2010,QZ_RMP2011}. It was proposed in a pioneering work by Haldane \cite{H_PRL1988} that a spinless fermionic model on a honeycomb lattice exhibits integer quantum Hall effect \cite{KDP_PRL1980} without an external magnetic field. This model, serving as the first example of Chern insulator, breaks the time reversal symmetry with a complex next-nearest-neighbor hopping and is characterized by a nonzero Chern number \cite{TKNN_PRL1982}. Later, the concept was generalized to time reversal symmetric systems with spin-orbital coupling (SOC), such as the graphene \cite{KM_PRL2005a,KM_PRL2005b} and HgTe/CdTe quantum wells \cite{BHZ_S2006,Ketc_S2007}. The SOC there generates complex hopping terms similar to that proposed by Haldane but with opposite chiralities for electrons with up spins and down spins, resulting in the quantum spin Hall insulator characterized by a $Z_2$ index.

@@

\par Another involved intriguing phenomenon arising from Coulomb repulsions in flat or nearly-flat bands is the itinerant ferromagnetism \cite{T_PTP1998}. It was proved by Tasaki and Mielke that the ground state of a flat electron band with its filling factor not more than but sufficiently close to $1/2$ is ferromagnetically ordered as long as an infinitesimal onsite Hubbard repulsion is present \cite{T_PRL1992,M_PLA1993,MT_CMP1993}. Afterwards, this ferromagnetism was shown to be stable against small nonflatness of the electron bands if and only if the Hubbard interaction exceeds a critical value \cite{T_PRL1994}. Spin wave excitations over this ferromagnetic ground state were also studied \cite{KA_PRL1994,SGDL_PRB2018} and itinerant topological magnons have been reported quite recently \cite{SGDL_PRB2018}.

\par The interplay between flatband ferromagnetism and nontrivial band topology enriches the related physics. In fact, ferromagnetism is essential in the generation of stable fractional Chern insulators in the proposals where the spin degrees of freedom of electrons are considered \cite{TMW_PRL2011,NSRCM_PRB2011,LBFL_PRL2012}. Furthermore, ferromagnetism can also lead to possible high-temperature quantum anomalous Hall effect (QAHE) when the nearly-flat topological band is half-filled \cite{NSRCM_PRL2012}. In this paper, we study the ferromagnetism and  spin excitations from the ferromagnetic ground state in nearly-flat topological bands. As a paradigm, we consider a square lattice Hubbard model whose free part is a $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). When the model is quarter filled (or correspondingly, the lower nearly-flat band is half filled), the ground state is spin fully polarized due to the ferromagnetism and exhibits QAHE because of the nonzero Chern number of a single-spin band. Consequently, the charge excitations are gapped and the low energy physics is dominated by the one-spin-flip excitations. These spin-1 excitations have been studied by a generalized bosonization scheme where the interacting fermionic model is mapped to a free bosonic model describing spin-wave excitations at the harmonic approximation \cite{DG_PRB2015}. The ferromagnetism was shown to be stable against such spin wave excitations, which are gapless in the Chern Hubbard model and gapped in the $Z_2$ Hubbard model. However in this bosonization scheme, the free part of the electron model plays no role in the spin wave excitations other than contributing a global constant, suggesting that it should fail due to the competition between the kinetic energy and potential energy of electrons when the nonflatness of the electron bands is not negligible \cite{T_PRL1994}. In fact, in a strictly local periodic tight-binding model, an energy band with a nonzero Chern number cannot be exactly flat \cite{CMST_JPMT2014}. Therefore it remains an open question on whether the ground state is stable against the spin-1 excitations and how the nonflatness of the electron bands manifests itself in the spin-1 excitation spectra in such models.

\par To elucidate these questions, we adopt the numerical exact diagonalization method with a projection onto the lower nearly-flat band to take close investigations on the spin-1 excitations of the models. A critical magnitude of the Hubbard interaction is found for both the Chern Hubbard model and the $Z_2$ Hubbard model, below which the ferromagnetic phase is unstable. Furthermore, the spin-1 excitation spectra are shown to consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which leads to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase. We also make a comparison of our results in the flatband limit with those obtained by the bosonization scheme \cite{DG_PRB2015}. Qualitative agreements are observed for the Chern Hubbard model, yet notable discrepancy appears for the $Z_2$ Hubbard model. We attribute the discrepancy to the self-interactions of the magnons, which are ignored in the bosonization scheme but treated exactly in our method after the projection.

\par The rest of the paper is organized as follows. In Sec. \ref{mm}, we introduce the Chern Hubbard model and $Z_2$ Hubbard model studied in this paper, discuss the nontrivial band topology of their free parts, interpret the emergence of QAHE resulting from the interplay between flatband ferromagnetism and nontrivial band topology, and formulate the exact diagonalization method with a projection onto the lower nearly-flat band on details. In Sec. \ref{nr}, we summarize the phase diagram and elaborate the spin-1 excitation spectra of both models. Section \ref{sd} provides a summary and discussion.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{lattice}
\caption{(Color online) (a) Schematic representation of the $\pi$-flux model. Blue and red solid circles denote the $A$ and $B$ sublattices, respectively. The nearest-neighbor hopping amplitudes (solid black lines) are equal to $t_{1}\exp(i\alpha_{\sigma}\pi/4)$ (see text) along the direction of the arrows, the next-nearest-neighbor hopping amplitudes are equal to $t_{2}$ (dashed green lines) and $-t_{2}$ (dashed purple lines). The shaded area denotes the unit cell. (b) First Brillouin Zone. $\Gamma=(0,0)$, $X_{1,2}=(\pm\pi,0)$, $Y_{1,2}=(0,\pm\pi)$, $M_{1,2,3,4}=(\pm\pi,\pm\pi)$, $O_{1,2,3,4}=(\pm\frac{\pi}{2},\pm\frac{\pi}{2})$.}
\label{lattice}
\end{figure}

\par We consider a generalized $\pi$-flux Hubbard model on the square lattice, whose Hamiltonian can be written as $H=H_0+H_U$, where $H_0$ is the spinfull genralization of the original spinless model proposed in Ref. \onlinecite{NSCM_PRL2011},
\begin{equation}
H_0=\sum_{\langle ij\rangle,\sigma}(t_1^{ij,\sigma}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.})
    +\sum_{\langle\langle ij\rangle\rangle,\sigma}(t_2^{ij}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.}),
\end{equation}
and $H_U$ is the Hubbard interaction
\begin{equation}
H_U=U\sum_{i}n_{i\uparrow}n_{i\downarrow}.
\end{equation}
Here, $c_{i\sigma}^{\dagger}(c_{i\sigma})$ creates (annihilates) a spin $\sigma$ electron at site $i$, $n_{i\sigma}=c^{\dag}_{i\sigma}c_{i\sigma}$ is the particle-number operator, $\langle ij\rangle$ denotes the nearest-neighbor (NN) bonds and $\langle\langle ij\rangle\rangle$ denotes the next-nearest-neighbor (NNN) bonds. As shown in Fig. \ref{lattice}(a), the spin-dependent NN hopping amplitude $t_1^{ij,\sigma}$ and the spin-independent NNN hopping amplitude $t_2^{ij}$ are given by
\begin{equation}
t_1^{ij,\sigma}=t_1\exp\left(i\delta_1^{ij}\alpha_\sigma\pi/4\right)
\end{equation}
and
\begin{equation}
t_2^{ij}=t_2\delta_2^{ij},
\end{equation}
respectively. Here, $\delta_1^{ij}=+1$ if the NN electron hopping is along the direction of the solid black arrow and $\delta_1^{ij}=-1$ if along the reversed direction. $\delta_2^{ij}=+1$ if the NNN electron hopping is along the dashed green lines and $\delta_2^{ij}=-1$ if along the dashed purple lines. The spin-dependent phase factor $\alpha_\sigma$ breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=\alpha_\downarrow=+1$, whereas it preserves the time-reversal symmetry but breaks the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=+1$ and $\alpha_\downarrow=-1$.

\par Due to the complex NN hopping, each electron will acquire a $\pi$ phase as it hops around a plaquette along the direction of the black arrows as indicated in Fig. \ref{lattice}(a). Therefore, $H_0$ describes free electrons hopping on a square lattice in the presence of a fictitious staggered $\pi$-flux pattern \cite{WWZ_PRB1989}. For the time-reversal-symmetry-breaking case, $\alpha_\uparrow=\alpha_\downarrow$, the fluxes experienced by spin-up electrons and spin-down electrons are the same, while for the time-reversal-symmetry-preserving case, $\alpha_\uparrow=-\alpha_\downarrow$, they are opposite.

\subsection{Topology of free term $H_0$}\label{topo_free_part}
\par Gapped noninteracting fermionic systems can be topologically classified by their Hamiltonians in the momentum space in the presence of symmetries \cite{SRFL_PRB2008}. After the Fourier transformation, the free part $H_0$ of our model reads

@@

\par When there is a gap between the $d$ band and $f$ band, these bands can be shown to be topologically non-trivial by calculating their Chern numbers \cite{TKNN_PRL1982} (for the time-reversal-symmetry-breaking case) or $Z_2$ indices \cite{KM_PRL2005b,SWSH_PRL2006} (for the time-reversal-symmetry-preserving case). The Chern number for a single spin component of the $d$ band or the $f$ band can be expressed in terms of the coefficients $D_{i,\mathbf{k}}$ \cite{HK_RMP2010,QZ_RMP2011},

@@

of the $d$ band is $1$ and nontrivial. As a consequence, the ground state of $H_0$ will be a noninteracting $Z_2$ insulator and exhibits quantum spin Hall effect(QSHE) when the lower $d$ band is fully filled.

\subsection{Emergence of QAHE in half-filled nearly-flat topological bands}\label{emerg_qahe}
\par For a free fermionic system hosting an energy band with a nonzero Chern number or $Z_2$ index, the distinguished phenomenon resulting from this nontrivial band topology, such as QAHE or QSHE, only manifests itself when the topological band is fully filled. At any fractional filling, the ground state of such a system will be a trivial metal. Intriguingly, when the Coulomb interactions between electrons is introduced, the physics of the nontrivial band topology becomes more involved, especially when the band is nearly flat so that the effects of the Coulomb interactions are highly enhanced. Combined with strong Coulomb interactions, nontrivial topological phases can emerge from fractionally filled topological bands. In this article, we are interested in half-filled strongly-correlated nearly-flat topological bands where QAHE can arise\cite{NSCM_PRL2011,DG_PRB2015}. The essence for the occurrence of this nontrivial phase is the emergence of itinerant ferromagnetism on nearly-flat bands\cite{T_PTP1998}, which fully polarizes all electron spins. Therefore, only one spin component of the topological band will be fully filled exactly, which leads to QAHE due to the nonzero Chern number of that spin component of the band.

\par The Chern Hubbard model and $Z_2$ Hubbard model on square lattice described above serve as the paradigm, where half filling of the lower electron band corresponds to quarter filling of the whole system because of the existence of $AB$ sublattices. When $t_2/t_1$ takes values in a selected region, the $d$ band and $f$ band are quite flat in that the flatness ratio $\Delta/W$, which is defined as the ratio of the gap $\Delta$ between these two bands to the bandwidth $W$ of the lower band, can be as large as $4.83$ \cite{NSCM_PRL2011,DG_PRB2015}. In the next subsection, we will introduce the exact diagonalization method with a projection onto the lower nearly-flat electron band to study the spin-1 excitations of the ferromagnetic phases in these two topological Hubbard models.

\subsection{Exact diagonalization with projection}\label{exact_diag_proj}
\par Exact diagonalization method with a projection onto the low-energy Hilbert space has been widely applied to systems that host flat or nearly-flat energy bands \cite{NSCM_PRL2011,RB_PRX2011,NSRCM_PRB2011,SGDL_PRB2018,NSRCM_PRL2012,DG_PRB2015}. This approach applies when the energy gap between the flat or nearly-flat band and other bands is larger than the Coulomb interaction.

@@

Let $|\text{FM}\rangle$ denotes the spin-up fully polarized state on the $d$ band,

@@

where FBZ denotes the first Brillouin zone [see Fig. \ref{lattice}(b)] and $|0\rangle$ is the fermion vaccum. Here, $d^{\dag}_{\mathbf{k}\uparrow}$ creates a spin-up electron with momentum $\mathbf{k}$. Then the basis of the spin-1 excitations with a center-of-mass momentum $\mathbf{q}$ over this reference state can be written as

@@

|\mathbf{k}_i\rangle_\mathbf{q}=d^{\dagger}_{\mathbf{k}_{i}-\mathbf{q}\downarrow}d_{\mathbf{k}_{i}\uparrow}|\text{FM}\rangle,

@@

which labels a spin-1 scattering channel with the index $\mathbf{k}_{i}$. The dimension of this Hilbert space scales linearly with respect to the number of electron momentums \cite{SGDL_PRB2018}, so a much larger system can be numerically accessed than the usual exact diagonalization without projection. It enables us to analyze the properties of the spin-1 excitation spectra in detail in the whole first Brillouin zone rather than some restricted discrete points solely. The matrix element of the projected Hamiltonian on this spin-1 excitation basis can be easily obtained after some algebra,

@@

Here $\delta_{\mathbf{k}_j,\mathbf{k}_i}$ is the Kronecker delta function. Then the full spin-1 excitation spectra can be obtained by the diagonalization of the matrix whose elements are defined by Eq. (\ref{PHP}). It is noted that $|\text{FM}\rangle$ is the true ground state only if the whole spin-1 excitation spectra have no negative energies. Thus we can use this as the criterion to determine the destabilization of the ferromagnetic phase.
\par We also want to give some remarks on the flatband limit in the framework of this method. The flatband-limit Hamiltonian shares the same single-particle eigenfunctions as well as the interaction terms with the original one, yet it has exactly-flat single-particle energy bands\cite{NSCM_PRL2011}. Thus the free part $h^\text{flat}_{\mathbf{k}\sigma}$ of the flatband-limit Hamiltonian can be defined as
